---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# geodiv

<!-- badges: start -->
<!-- badges: end -->

*geodiv* calculates gradient surface metrics in R. These metrics are applied to continuous spatial data (i.e., rasters or matrices) and represent spatial heterogeneity. 

## Installation

You can install the released version of *geodiv* from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("geodiv")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")

# install from master branch
devtools::install_github("bioXgeo/geodiv")

# install from development branch
devtools::install_github("bioXgeo/geodiv",ref = "dev")
```
On Mac OS X, you may need to install the development tools here to get the package to install:

<https://cran.r-project.org/bin/macosx/tools/>

## Example

This is a basic example which shows you how to calculate several metrics over an entire image. *geodiv* may also be applied with moving windows over an entire image using the 'texture_image' function. For a more complex example, see the vignette.

```{r example}
library(geodiv)

# import example raster
data(normforest)

# apply metrics
sa(normforest) # average surface roughness
svk(normforest) # reduced valley depth
ssc(normforest) # mean summit curvature

```
